Hydras \& Co.: Formalized mathematics in Coq\\ for inspiration and entertainment

### 2022-05-11 doc/JFLA2022/Slides/Slides.tex
work on JFLA Slides

  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\centering
\scalebox{0.7}{
\begin{tikzpicture}[thick]

@@

\begin{scope}[xshift=-5.7cm]
\node[rectangle, dotted,draw=black,minimum height=0.5cm,minimum width=1.5cm] (goedel) { Goedel\footnote{R. O'Connor} };
\node[rectangle, dotted,draw=black,minimum height=0.5cm,minimum width=1.5cm,left of=goedel, node distance=2.2cm] (pock) { Pocklington };
\node[rectangle, dotted,draw=black,minimum height=0.5cm,minimum width=1.5cm,right of=goedel, node distance=2.5cm] (cantor) { Cantor\footnote{C. \& E. Contejean}};

\node[rectangle, dotted,draw=black,minimum height=0.5cm,minimum width=1.5cm,right of=cantor, node distance=3.6cm] (cats) { Categories in ZFC };
\node[rectangle, dotted,draw=black,minimum height=0.5cm,minimum width=1.5cm,right of=cats, node distance=3.2cm] (additions) {Additions};

@@

\begin{scope}[yshift=-0.8cm,xshift=-1.45cm]
\node[rectangle, draw=black,minimum height=0.5cm,minimum width=1.5cm] (mathcomp) { MathComp };

@@

\draw[lookcolor,->,dotted] (acl2.south) -- (cantor.north) ;
\draw[lookcolor,->,dashed] (additions.south) -- (chains.north) ;
\draw[lookcolor,->,dashed] (goedel) -- (goedelcc) ;
\draw[lookcolor,->,dashed] (goedel) -- (hydras.north) ;
\draw[lookcolor,->,dashed] (cantor) -- (hydras.north east) ;
\draw[lookcolor,->,dashed] (cantor.south) -- (gaia.north west) ;
\draw[lookcolor,->,dashed] (pock) -- (pockcc) ;
\draw[lookcolor,->,dashed] ([xshift=2.1cm]cats) -- (gaia) ;

@@

\pause
\textbf{Remark:} The proof is not trivial. The built path could miss $\beta$!

@@

\begin{frame}[fragile]

      \begin{block}{}
    The functions \textcolor{mathcolor}{$H'_\alpha, F_\alpha\;(\alpha< \epsilon_0)$} are implemented with \textcolor{plugincolor}{coq-equations}.
    \vspace{5pt}
    \begin{footnotesize}
      \inputsnippets{../Hprime_HprimeDef}   
    \end{footnotesize}

@@

  \begin{block}{Theorem}
    The function which computes the length of hydra battles
    (according to the initial index) is not primitive recursive
    (for $\alpha\geq \omega^\omega$).
    \vspace{4pt}
  \end{block}

  \begin{block}{Sketch of proof}

    We prove and apply a famous lemma:

       \begin{quote}

      {\color{mathcolor}
        \[
          \exists q, \forall x_1,\dots,x_n, 
          f(x_1,\dots,\,x_n)\leq\textrm{Ack}(q,\textrm{max}(x_1,\dots,x_n))
        \]}
    \end{quote}

@@

% \begin{frame}

%   \begin{block}{}
%     {\color{coqstylecolor}
%       \begin{itemize}
%       \item \emph{In order to prove this theorem, we chose to host and  maintain under \community Russel O'Connor's library \textcolor{plugincolor}{goedel}, which contains a formalization of primitive recursive functions.}

%       \item    \emph{The interest of O'Connor's contribution led us to write a chapter on primitive recursive functions, with [counter-]examples and exercices.}

%       \item \emph{Proving formally that a given function is not primitive recursive is a nice example of application of a complex (mutually recursive, dependently typed) induction principle. You have to take into account the set of PR functions of any arity.}

%       \end{itemize}
%     }
%   \end{block}
% \end{frame}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Project Organization}
\begin{frame}
  \frametitle{Project  Organization}
  \begin{itemize}
   \item Compatibility with \gaia
  \item Integration with \community
  \item Documentation with \alectr
  \item Package dependencies
  \item Continuous integration
  \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{Compatibility with Gaia}
  \begin{block}{From \Hydras to \gaia}
    {\footnotesize
      \inputsnippets{./canonDef}
      \inputsnippets{./gcanonLimitOf}}
  \end{block}
\end{frame}

@@

  \begin{block}{}
  {\color{lookcolor}  With \Hydras, we hope to demonstrate that projects with complex dependencies are feasible to manage using recent advances in build management and infrastructure.}
  \end{block}

@@

  \begin{block}{To do ?}
    Define a language for
    \begin{itemize}
    \item Searching sentences in any \coq file,
    \item Applying filters in order to generate as many snippets as needed.
    \end{itemize}
  \end{block}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% \begin{frame}
%   \frametitle{Package organization and dependencies}

% \begin{figure}[ht]
% \centering
% \scalebox{0.7}{
% \begin{tikzpicture}[thick]

% \begin{scope}[xshift=-6cm]
% \node[rectangle, dotted,draw=black,minimum height=0.5cm,minimum width=1.5cm] (goedel) { Goedel };
% \node[rectangle, dotted,draw=black,minimum height=0.5cm,minimum width=1.5cm,left of=goedel, node distance=2.2cm] (pock) { Pocklington };
% \node[rectangle, dotted,draw=black,minimum height=0.5cm,minimum width=1.5cm,right of=goedel, node distance=2.5cm] (additions) { Additions };
% \node[rectangle, dotted,draw=black,minimum height=0.5cm,minimum width=1.5cm,right of=additions, node distance=2cm] (cantor) { Cantor };

% \node[rectangle, dotted,draw=black,minimum height=0.5cm,minimum width=1.5cm,right of=cantor, node distance=3.2cm] (cats) { Categories in ZFC };
% \end{scope}

% \begin{scope}[yshift=-1cm,xshift=0.45cm]
%   \node[rectangle, draw=black,fill=blue!30,minimum height=0.5cm,minimum width=1.5cm] (paramcoq) { Paramcoq };
% \node[rectangle, draw=black,minimum height=0.5cm,minimum width=1.5cm,right of=paramcoq, node distance=2.5cm] (mathcomp) { MathComp };
% \node[rectangle, draw=black,minimum height=0.5cm,minimum width=1.5cm,node distance=2cm,left of=paramcoq, node distance=4.5cm] (equations) { Equations };
% \end{scope}

% \begin{scope}[yshift=-2cm,xshift=-4cm]
% \node[rectangle, draw=black,fill=orange!80,minimum height=0.5cm,minimum width=1.5cm] (hydras) { Hydra-battles };
% \node[rectangle, draw=black,fill=blue!30,minimum height=0.5cm,minimum width=1.5cm,left of=hydras, node distance=3.5cm] (pockcc) { Pocklington };
% \node[rectangle, draw=black,fill=orange!80,minimum height=0.5cm,minimum width=1.5cm,right of=hydras, node distance=2.5cm] (chains) { Addition-chains };
% \node[rectangle, draw=black,fill=blue!30,minimum height=0.5cm,minimum width=1.5cm,right of=chains, node distance=2.2cm] (gaia) { Gaia };
% \end{scope}

% \begin{scope}[yshift=-3cm,xshift=-6cm]
%  \node[rectangle, draw=black,fill=blue!30,minimum height=0.5cm,minimum width=1.5cm] (goedelcc) { Goedel };
%  \node[rectangle, draw=black,fill=orange!80,minimum height=0.5cm,minimum width=1.5cm,right of=goedelcc, node distance=3cm] (gaiahydras) { Gaia-hydras };
% \end{scope}

% \draw[lookcolor,->,dashed] (additions.south) -- (chains.north) ;
% \draw[lookcolor,->,dashed] (goedel) -- (goedelcc) ;
% \draw[lookcolor,->,dashed] (goedel) -- (hydras.north west) ;
% \draw[lookcolor,->,dashed] (cantor) -- (hydras.north east) ;
% \draw[lookcolor,->,dashed] (cantor.south) -- (gaia.north west) ;
% \draw[lookcolor,->,dashed] (pock) -- (pockcc) ;
% \draw[lookcolor,->,dashed] ([xshift=2.1cm]cats) -- (gaia) ;

% \draw[->] (hydras) -- (gaiahydras) ;
% \draw[->] (hydras.south west) -- (goedelcc) ;
% \draw[->] (gaia.south west) -- (gaiahydras.east) ;

% \draw[->] (equations) -- (hydras) ;
% \draw[->] (mathcomp) -- (gaia) ;
% \draw[->] (mathcomp) -- (chains) ;
% \draw[->] (paramcoq.south) -- (chains) ;
% \draw[->] (pockcc) -- (goedelcc) ;
% \end{tikzpicture}
% }
% \caption{\textcolor{lookcolor}{Genealogy} and dependencies for \Hydras packages.}
% Dotted boxes represent historical Coq contributions, while regular boxes represent maintained \coq packages. Orange packages are maintained in the \Hydras GitHub repository, while blue packages are maintained in other \community repositories. Dotted lines represent \coq code ancestry, while regular lines represent direct code dependencies.}

%   \label{fig:genealogy}
% \end{figure}

% %%%% TODO : Resize  !
% {\scriptsize
% \begin{table}[ht]
% \centering
% \scriptsize
% \begin{tabular}{llllrr}
% \hline
% \textbf{Package name} &  \textbf{Version} & \textbf{Spec LOC} & \textbf{Proof LOC}\\
% \hline
% Pocklington & \texttt{8.12.0} & 825 & 3,798 \\
% Goedel  & \texttt{8.13.0} & 2,799 & 10,762 \\
% Gaia &  \texttt{1.12} & 28,850 & 124,839 \\
% Hydra-battles & \texttt{0.5} & 14,368 & 48,385 \\
% Addition-chains & \texttt{0.5} & 1,961 & 2,210 \\
% Gaia-hydras &  \texttt{0.8} & 677 & 2,023 \\
% \hline
% \end{tabular}
% \caption{\scriptsize Current numbers of lines of code for packages in the \Hydras galaxy.}
% \label{tbl:loc}
% \end{table}}
%\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
  \frametitle{\Hydras: genealogy and dependencies}
\centering
\scalebox{0.6}{
\begin{tikzpicture}[thick]

@@

\begin{scope}[xshift=-5.7cm]
\node[rectangle, dotted,draw=black,minimum height=0.5cm,minimum width=1.5cm] (goedel) { Goedel\footnote{R. O'Connor} };
\node[rectangle, dotted,draw=black,minimum height=0.5cm,minimum width=1.5cm,left of=goedel, node distance=2.2cm] (pock) { Pocklington\footnote{O. Caprotti \& M. Oostdijk}};
\node[rectangle, dotted,draw=black,minimum height=0.5cm,minimum width=1.5cm,right of=goedel, node distance=2.5cm] (cantor) { Cantor\footnote{C. \& E. Contejean}};

\node[rectangle, dotted,draw=black,minimum height=0.5cm,minimum width=1.5cm,right of=cantor, node distance=3.6cm] (cats) { Categories in ZFC\footnote{C. Simpson} };
\node[rectangle, dotted,draw=black,minimum height=0.5cm,minimum width=1.5cm,right of=cats, node distance=3.2cm] (additions) {Additions};

@@

\begin{scope}[yshift=-0.8cm,xshift=-1.2cm]
\node[rectangle, draw=black,minimum height=0.5cm,minimum width=1.5cm] (mathcomp) { MathComp };

@@

\draw[lookcolor,->,dotted] (acl2.south) -- (cantor.north) ;
\draw[lookcolor,->,dashed] (additions.south west) -- (chains.north) ;
\draw[lookcolor,->,dashed] (goedel) -- (goedelcc) ;
\draw[lookcolor,->,dashed] (goedel) -- (hydras.north) ;
\draw[lookcolor,->,dashed] (cantor) -- (hydras.north east) ;
\draw[lookcolor,->,dashed] (cantor.south) -- (gaia.north west) ;
\draw[lookcolor,->,dashed] (pock) -- (pockcc) ;
\draw[lookcolor,->,dashed] ([xshift=2.1cm]cats) -- (gaia) ;

@@

%\pause
\textbf{Remark:} The proof is not trivial. The built path could miss $\beta$!

@@

% \begin{frame}[fragile]

%       \begin{block}{}
%     The functions \textcolor{mathcolor}{$H'_\alpha, F_\alpha\;(\alpha< \epsilon_0)$} are implemented with \textcolor{plugincolor}{coq-equations}.
%     \vspace{5pt}
%     \begin{footnotesize}
%       \inputsnippets{../Hprime_HprimeDef}   
%     \end{footnotesize}

%   \end{block}

@@

  \begin{block}{Theorems}
    \begin{itemize}
    \item $F_\alpha$ is primitive recursive iff $\alpha<\omega$.
     \item  
    The function which computes the length of hydra battles
    for $\alpha$ (according to the initial index) is not primitive recursive
    (for $\alpha\geq \omega^\omega$).

    \end{itemize}

  \begin{block}{Sketch of proof:}

   We apply monotony properties and the famous lemma:
    \vspace{5pt}

       \begin{quote}
      {\color{mathcolor}
        For any $n$-ary primitive recursive  function $f$, 
        there exists some $q$ such that the following inequality holds:
         \[\forall\,x_1,\dots,x_n,  
          f(x_1,\dots,\,x_n)\leq\textrm{Ack}(q,\textrm{max}(x_1,\dots,x_n))\]}
    \end{quote}
    (Proof by induction on any primitive recursive function of any arity.)
  \end{block}

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Project Organization}
% \begin{frame}
%   \frametitle{Project  Organization}
%   \begin{itemize}
%    \item Compatibility with \gaia
%   \item Integration with \community
%   \item Documentation with \alectr
%   \item Package dependencies
%   \item Continuous integration
%   \end{itemize}
% \end{frame}

@@

\begin{frame}
  \frametitle{Compatibility with Gaia}
  \begin{block}{From \Hydras to \gaia}
    {\footnotesize
      \inputsnippets{./canonDef}
      \inputsnippets{./gcanonLimitOf}}
  \end{block}
  \begin{block}{}
    We experiment also the other direction (importing \texttt{ssete9} lemmas to \Hydras).
  \end{block}
\end{frame}

@@

  \begin{block}{}
    With \Hydras, we hope to demonstrate that projects with complex dependencies are feasible to manage using recent advances in build management and infrastructure.
  \end{block}

@@

  \begin{block}{To do}
    Define a language for
    \begin{itemize}
    \item Searching sentences in any \coq file,
    \item Applying filters in order to generate as many snippets as needed.
    \end{itemize}
  \end{block}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
